Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-04-03 Publications/2015-NeuroImage/03-introduction.tex
refactoring, add figure of phantoms

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely and coregistration of anatomical
  images are difficult for several reasons.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  interfacing voxels due to the low resolution of images (generally around
  \isores{2.2} [mm]).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  use for direct segmentation.
The contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Finally, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of presenting geometrical distortion and 
  signal losses \citep{jezzard_correction_1995} that hinder the segmentation by
  registering an anatomical image as source of same-subject prior information.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, there is a large variability in the mask obtained depending on the value of
  choice \citep{taoka_fractional_2009}.
The approach is unreliable for the high variability of \gls*{fa} and its
  decay with the increase of incoherence between fibers.
\cite{zhukov_level_2003} proposed active contours with edges represented
  by level sets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based level-sets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targets the definition of appropriate features to segmentation,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  and graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely and coregistration of anatomical
  images are difficult for several reasons.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  interfacing voxels due to the low resolution of images (generally around
  \isores[mm]{2.2}).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  use for direct segmentation.
The contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Finally, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of presenting geometrical distortion and 
  signal losses \citep{jezzard_correction_1995} that hinder the segmentation by
  registering an anatomical image as source of same-subject prior information.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri} \citep{irfanoglu_effects_2012}, particularly in certain fiber bundles
  when used in whole-brain tractography.

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, the mask and subsequent analyses strongly depend on the threshold set
  \citep{taoka_fractional_2009} owing to the large variability of \gls*{fa} and its
  decay with the increase of incoherence between fibers.
\cite{zhukov_level_2003} proposed active contours with edges represented
  by level sets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based level-sets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targets the definition of appropriate features to segmentation,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  and graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

### 2015-04-03 Publications/2015-NeuroImage/04-methods.tex
reworking intro and methods
refactoring, add figure of phantoms

  \begin{align}
  U_s\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
The displacements along the \gls*{pe} axis are related to the local field
  deviation $\Delta B_0(\vec{r})$ \citep{jezzard_correction_1995}:

% Joint segmentation and distortion correction of \gls*{dmri} data is also interesting to
%   improve the overall consistence of complex processing workflows.
% The closest antecedent of \emph{regseg} is \emph{bbregister} \citep{greve_accurate_2009}.
% Both methods use active contours to align surfaces to same-subject \gls*{dmri} images.
% However, \emph{bbregister} is limited to affine transformations, and thus, it is not able to
%   solve the susceptibility-derived distortions, generally nonlinear.
% To enable registration of uncorrected data, \emph{bbregister} excludes from the cost-function
%   those regions frequently warped.
% Conversely, \emph{regseg} implements a free-deformation field, supported with B-Spline
%   functions to estimate deformations.
% A second diverging foundation is the active-contours framework.
% In \emph{regseg}, active contours are implemented without edges \citep{chan_active_2001},
%   enabling the tool to segment homogeneous regions within the brain.
% On the other hand, \emph{bbregister} requires that the target features present intensity
%   boundaries as it implements active contours with edges.
% The tissue interfaces imaged with \gls*{dmri} are affected by partial volume effects that
%   blur boundaries.
% In practice, registration of \gls*{t1} and \gls*{dmri} data with \emph{bbregister}
%   is limited to the linear match of the pial surface of the brain.

% Our method is close to \citep{guyader_combined_2011} and \citep{gorthi_active_2011}
%   in the use of active contours without edges for registration, however we avoided
%   the use of level sets with the application of shape gradients
%   \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
% \cite{gorthi_active_2011} used a gaussian regularizer, and \cite{guyader_combined_2011}
%   proposed a nonlinear elasticity smoother.
% In contrast, \emph{regseg} implements the anisotropic regularization proposed by
%   \citep{nagel_investigation_1986}.
% We demonstrate its aptness on the correction of the nonlinear distortions typically
%   presented by \gls*{dmri} data of the brain.
% Therefore, a typical protocol for the extraction of structural connectivity
%   comprehends \gls*{dmri}, \gls*{t1}, and the acquisition determined by the correction
%   method of choice \citep{daducci_connectome_2012}.

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each phantom type, with two resolutions each
  (1.0mm and 2.0mm isotropic).
Field of view at both resolutions was \isores{100}mm.
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted at high resolution, using 
  \texttt{mri\_tessellate}\footnote{A marching-cubes algorithm shipped with 
  \emph{FreeSurfer}.}.

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03a) at (0,14.8)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-a}};
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,17.3) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{B}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{C}};
      \end{tikzpicture}
    }%
	\caption{A. The ``cortex'' phantom is a spherical shape with two sulci and an
	  outer crust resembling the cortical folding (left).
	The model is used to generate \gls*{t1} and \gls*{t2} images after warping the
	  contours using a random and plausible transformation $U_{true}^{-1}$ (right).
	B. Visual assessment of the results on the low resolution sets:
	  ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
	  and ``box'' at (bottom-right).
	In yellow color, the recovered contours after registration are represented.
	Our method showed high accuracy, as demonstrates almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
	Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
	  clustering methods, but it is successfully segmented with our method.
	C. Quantitative evaluation of registration error in terms of average Hausdorff distance of
	  surfaces at high (left) and low (right) resolutions, demonstrating that the error is
	  consistently below the voxel size.
	  }\label{fig:phantom}
\end{figure*}

@@

>>>>

  \begin{align}
  U_s\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+\vec{u}(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = \vec{u}(\vec{r})$ is the displacement of every point with respect
  to the reference domain.

@@

\paragraph*{Simulated phantoms}%
\label{sec:digital_phantoms}
We proved the concept on simplistic digital phantoms which we will name after their
  appearance as ``box'', ``ball'', ``L-shape'', and ``gyrus'' (\autoref{fig:phantom},
  boxes A and B).
Using \emph{phantomas} \citep{caruyer_phantomas_2014}, we simulated
  \gls*{t1} (TE/TR=10/1500ms) and \gls*{t2} images (TE/TR=90/5000ms)
  corresponding to each phantom type, with two resolutions each
  (1.0mm and 2.0mm isotropic).
The field of view at both resolutions was \isores[mm]{100}.
Simulations were corrupted with rician noise for a \gls*{snr} of 300.0.
The reference surfaces were extracted at high resolution, using 
  \texttt{mri\_tessellate}\footnote{A marching-cubes algorithm shipped with 
  \emph{FreeSurfer}.}.

\begin{figure*}
\includegraphics[width=\linewidth]{figures/figure03}
\caption{Evaluation workflow of phantoms}\label{fig:evphantoms}
\end{figure*}

@@

The displacements along the \gls*{pe} axis are related to the local field
  deviation $\Delta B_0(\vec{r})$ \citep{jezzard_correction_1995}:

### 2015-04-03 Publications/2015-NeuroImage/05-results.tex
fix aggregate, update results
refactoring, add figure of phantoms

\subsection{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure04.pdf}
	\caption{Assessment of the segmentation model, sampling the value of the energy functional
	\eqref{eq:energy} versus several registration errors generated as explained in
	\autoref{sec:res_model_and_metric}.
	The metric of our registration method displayed a smooth gradient towards the minimum
	located at the $\epsilon = 0.0$ point.}\label{fig:energymap}
\end{figure}

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we compared the performance of \emph{regseg} and the standard \gls*{t2b}
  correction.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after correction,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses of whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregated CI for \emph{regseg} was $[1.08 - 1.50] mm$, whereas the competing method
  yielded an aggregated CI of $[2.06 - 2.43] mm$.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  violin plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is included in the \suppl{section S5}.

\begin{table}
		\centering
		\footnotesize
		\tabcolsep=0.05cm
    \begin{tabular}{cccccc}
    & & $\Gamma_{VdGM}$  & $\Gamma_{WM}$ & $\Gamma_{pial}$ & Aggregated \\
    \hline
    \multirow{2}{*}{CI}
       & \emph{regseg}        & 0.77 - 1.18 & 0.95 - 1.85 & 1.33 - 2.13 & 1.08 - 1.50 \\
       & T2B                  & 1.73 - 2.62 & 1.96 - 2.40 & 2.16 - 2.63 & 2.06 - 2.43 \\
    \hline
    \multirow{2}{*}{ANOVA}
       & p-value  & 1.4$\times$10$^{-5}$& 1.3$\times$10$^{-3}$& 2.5$\times$10$^{-3}$ & 2$\times$10$^{-9}$ \\
       & f-stat   & 27.87               & 12.89               & 11.13                & 45.04 \\
    \hline
    \end{tabular}
    \caption{All the 95\% CIs of the \gls*{swindex} distributions computed for the
      surfaces of interest were lower for \emph{regseg} with respect to the competing
      method.
    CIs reported in this table were computed using bootstrapping with mean as location
      statistic and $10^4$ samples.
    One-way-ANOVA tests indicated that there is a significant difference between our method and
      the competing method of choice.
    }\label{tab:results_real}
\end{table}

>>>>

\subsection{Proof of concept on digital phantoms}
\label{sec:results_phantom}

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{B}};
      \end{tikzpicture}
    }%
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as demonstrates almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at high (left) and low (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}

Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

@@

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we compared the performance of \emph{regseg} and the standard \gls*{t2b}
  correction.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after correction,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses of whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregate CI for \emph{regseg} was $[1.08 - 1.50] mm$, whereas the competing method
  yielded an aggregate CI of $[2.06 - 2.43] mm$.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  violin plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is included in the \suppl{section S5}.

\begin{table}
		\centering
		\footnotesize
		\tabcolsep=0.05cm
    \begin{tabular}{cccccc}
    & & $\Gamma_{VdGM}$  & $\Gamma_{WM}$ & $\Gamma_{pial}$ & Aggregate \\
    \hline
    \multirow{2}{*}{CI}
       & \emph{regseg}        & 0.50 - 0.78 & 0.50 - 0.55 & 0.66 - 0.73 & 0.56 - 0.66 \\
       & T2B                  & 1.78 - 2.58 & 1.94 - 2.36 & 2.16 - 2.58 & 2.05 - 2.39 \\
    \hline
    \multirow{2}{*}{ANOVA}
       & p-value  & 1.5$\times$10$^{-8}$& 6.3$\times$10$^{-16}$& 4.2$\times$10$^{-16}$ & 4$\times$10$^{-34}$ \\
       & f-stat   & 60.03               & 255.82               & 263.97                & 377.82              \\
    \hline
    \end{tabular}
    \caption{All the 95\% CIs of the \gls*{swindex} distributions computed for the
      surfaces of interest were lower for \emph{regseg} with respect to the competing
      method.
    CIs reported in this table were computed using bootstrapping with mean as location
      statistic and $10^4$ samples.
    One-way-ANOVA tests indicated that there is a significant difference between our method and
      the competing method of choice.
    }\label{tab:results_real}
\end{table}

### 2015-04-03 Publications/2015-NeuroImage/supplemental.tex
refactoring, add figure of phantoms

>>>>

\subsection{Assessment of the segmentation model}
\begin{figure}
  \includegraphics[width=\linewidth]{figures/figure04.pdf}
  \caption{Assessment of the segmentation model, sampling the value of the energy functional
  \eqref{eq:energy} versus several registration errors generated as explained in
  \autoref{sec:res_model_and_metric}.
  The metric of our registration method displayed a smooth gradient towards the minimum
  located at the $\epsilon = 0.0$ point.}\label{fig:energymap}
\end{figure}
